---
title: "Genetic algorithm using the evola package"
author: "Giovanny Covarrubias-Pazaran"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Genetic algorithm using the evola package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

The evola package is nice wrapper of the AlphaSimR package that enables the use of the evolutionary algorithm to solve complex questions in a simple manner.

The vignettes aim to provide several examples in how to use the evola package under different optimization scenarios. We will spend the rest of the space providing examples for: 

1) Optimizing the selection of one feature with a constraint in other
2) Obtaining optimal N individuals from a population for a given feature
3) Obtaining optimal N crosses from a population for a given feature
4) Optimizing a sample of size N to be representative
5) Optimizing a sample of size N to be predict another sample
6) How to specify constraints:
5a) gender
6b) number of times a parent should be used
7) How to optimize the number of progeny to produce per cross

### 1) Optimizing the selection of one feature with a constraint in other

The example presented here is a list of gems (Color) that have different weights in Kg (Weight) and a given value (Value). 

```{r}
library(evola)
set.seed(1)
# Data
Gems <- data.frame(
  Color = c("Red", "Blue", "Purple", "Orange",
            "Green", "Pink", "White", "Black", 
            "Yellow"),
  Weight = round(runif(9,0.5,5),2),
  Value = round(abs(rnorm(9,0,5))+0.5,2)
)
head(Gems)
```

The task to optimize here is to be able to pick in your bag all the possible gems that maximize the value with the contraint that your bag would break after 10Kg. In the evolafit function this would be specified as follows:

```{r}
# Task: Gem selection. 
# Aim: Get highest combined value.
# Restriction: Max weight of the gem combined = 10. 
res0<-evolafit(cbind(Weight,Value)~Color, dt= Gems,
               # constraints: if greater than this ignore
               constraintsUB = c(10,Inf), 
               # constraints: if smaller than this ignore
               constraintsLB= c(-Inf,-Inf), 
               # weight the traits for the selection
               traitWeight = c(0,1), 
               # population parameters
               nCrosses = 100, nProgeny = 20, recombGens = 1, 
               # coancestry parameters
               A=NULL, lambda=c(0,0), nQTLperInd = 1, 
               # selection parameters
               propSelBetween = .9, propSelWithin =0.9, 
               nGenerations = 50
) 
```

Noticed that the formula cbind(Weight,Value)~Color specified the traits to be considered in the optimization problem are specified in the left side of the formula whereas the right side of the formula specifies the term corresponding to the genes that will form the 'genome' of the solution. Each trait in the formula requires a value for the contraints, weights in the fitness function (i.e., selection index) and lambda (weight for group relationship between the genes in the genome). The rest of the parameters are the parameters controling the evolution of the population of solutions.

When looking at the results of the evolution we can observe that the best solution for the traits under the contraints can be extracted with the bestSol() function.

```{r}
best=bestSol(res0)[1] # best solution for trait 1
xa = res0$M[best,] %*% as.matrix(Gems[,c("Weight","Value")]); xa
res0$M[best,]
```

The best selection of Gems was the one one found in the M element of the resulting object.

### 2) Obtaining optimal N individuals from a population for a given feature

TBD

```{r, fig.show='hold'}
# TBD
```

### 3) Obtaining optimal N crosses from a population for a given feature

TBD

```{r}
#TBD
```

### 4) Optimizing a sample of size N to be representative

TBD

```{r}
# TBD
```

### 5) Optimizing a sample of size N to be predict another sample

TBD

```{r}
# TBD
```

### 6) How to specify constraints

#### Gender
In this case is better if you only create the cross combinations that are possible (e.g., where male and female can couple) and you handed them to the evolutionary algorithm. That means, the rows of the crosses to be in the searching space only include the realistic ones.

#### Number of times a parent should be used

In this case you would need to do 2 things:

A) Create two new column with the mother and father of the cross
B) Create a new trait that is computed on the fly while runing the analysis where the trait is a representation of how many times the parent is present in the current contribution. For example the trait function can be defined for mother as:

```{r}
# data$motherN <- as.numeric(as.factor(data$mother))
# data$fatherN <- as.numeric(as.factor(data$father))
# fitnessf <- list(motherN= function(motherN,fatherN){
# v = table(c(motherN,fatherN))
# res <- ifelse(any(v > 4), 0, 1)
# return(res)
# })
```

Where motherN is a new trait that is just a numerical recoding of the mothers (e.g., as.numeric(as.factor(dt$mother)) ). This new fitness function for the trait motherN can be added to the 'fitnessf' argument together with a constraint (e.g., constraintsLB=1) to ensure that only crosses that meet the number of times showing in a cross are among the selected candidates.

### 7) How to optimize the number of progeny to produce per cross

The advice here is to upload directly the phased genotypes (haplotypes) to the AlphaSimR machinery and simulate the possible crosses to explore how many individuals are required to sample a given trait (oligogenic or polygenic) with a given probablility. No need to use the evola package.

## Literature

Giovanny Covarrubias-Pazaran (2024).  evola: a simple evolutionary algorithm for complex problems. To be submitted to Bioinformatics.

Gaynor, R. Chris, Gregor Gorjanc, and John M. Hickey. 2021. AlphaSimR: an R package for breeding program simulations. G3 Gene|Genomes|Genetics 11(2):jkaa017. https://doi.org/10.1093/g3journal/jkaa017.

Chen GK, Marjoram P, Wall JD (2009). Fast and Flexible Simulation of DNA Sequence Data. Genome Research, 19, 136-142. http://genome.cshlp.org/content/19/1/136.
